### 2018-05-26 background.tex

In this blocked algorithm, the loops over the $N$, $K$ and $M$
matrix dimensions are tiled by sizes $N_c$, $K_c$ and $M_c$ at 
layers 1, 2 and 3, respectively.
At layers 4 and 5, the $N$ and $M$ dimensions are further tiled by sizes
$N_r$ and $M_r$, respectively.
As a result, the innermost loop at layer 6 goes over the $K$
dimension for a total of $K_c$ times,
with each iteration performing a rank-1 update on
the $M_r \times N_r$ submatrix of $C$, as shown at layer 7.
In OpenBLAS, matrix $C$ is scaled by $\beta$ before the loops,
so the multiply-accumulate operation at layer 7 does not have to scale $C$ again.
Tiling factors $N_c$, $K_c$, $M_c$, $N_r$ and $M_r$ are carefully selected so that
matrices on each level fit into a certain level in the memory hierarchy.
Equations~(\ref{eq:constraints.reg}) -- (\ref{eq:constraints.l3}) express
the constraints on these tiling factors:
\begin{eqnarray}
  es (M_r + N_r + M_r N_r) & \le & c_{0} \label{eq:constraints.reg}\\
  nt_{1} \cdot es (N_r K_c + 2 M_r K_c) & \le & c_{1} \label{eq:constraints.l1}\\
  nt_{2} \cdot es (M_c K_c + 2 N_r K_c) & \le & c_{2} \label{eq:constraints.l2}\\
  es (N_c K_c + nt_{3} \cdot M_c K_c)   & \le & c_{3} \label{eq:constraints.l3}
\end{eqnarray}
Here,
$es$ denotes the size of a matrix element, e.g.,
8B for a double precision floating-point number,
and $c_l$ denotes the capacity of a register or cache 
at level $l$.
By (\ref{eq:constraints.reg}), $M_r$ and $N_r$ are so
constrained that $M_r$ elements from $A$,
$N_r$ elements from $B$ and $M_r \times N_r$ elements from $C$ can fit into the registers
(the pseudo level-0 cache).
By (\ref{eq:constraints.l1}), $K_c$ is so
constrained that $B_4$ ($N_r \times K_c$),
$A_4$ ($M_r \times K_c$), and $A_4$'s counterpart in the next iteration
at layer 5 fit into the L1 cache.
By (\ref{eq:constraints.l2}), $M_c$ is so
constrained that $A_3$ ($M_c \times K_c$),
$B_3$ ($N_r \times K_c$), and $B_3$'s counterpart in the next iteration
at layer 4 fit into the L2 cache.
Finally, by (\ref{eq:constraints.l3}), $N_c$ is so
constrained that 
$B_2$ ($K_c \times N_c$) and $A_2$ ($M_c \times K_c$) fit into the L3 cache.
Larger matrices at higher layers 1 -- 3 always reside in main memory.

@@

As stated in Section~\ref{sec:intro}, a GEMM implementation
is made up by a kernel routine and an overall workload-partitioning strategy.
In Figure~\ref{fig:gemm}, GEBP serves as the kernel routine,
and the overall workload-partitioning strategy is defined by the three outer loops at layers 1 -- 3.
Developers can decide how to partition the workload by choosing
tile factors $M_c$, $N_c$ and $K_c$,
and how to schedule GEBP tasks by interchanging loop orders at layers 1 - 3.
Either the N loop at layer 1 or the M loop at layer 3, or 
both,
can be parallelized in a multi-threaded context.
Specifically, the implementation in Figure~\ref{fig:gemm}
chooses an NKM loop order and parallelizes the M loop at layer 3.
Different threads work on different parts of $A_1$
and share the same $B_1$.
As a result, there are multiple $A_2$ instances
(one per thread)
and a single $B_2$ instance (shared by all threads).
To understand how the loop at layer 3 is parallelized on the 4-core processor, 
Figure~\ref{fig:workload} shows the data accessing patterns of a single thread $T_1$.
While all color shaded submatrices are accessed by $T_1$,
only the z-curve masked submatrices (a subset of color shaded ones)
are packed by the thread.
In other words, each thread packs its own $A_2$ instance and  all the threads collaboratively pack the 
unique
$B_2$ instance.
%% We will use the OpenBLAS GEMM implementation in Figure.~\ref{fig:gemm}
%% in discussion and evaluation throughout the paper.

@@

In contrary to intra-thread cache conflicts,
there exists another kind of cache conflicts, called
inter-thread cache conflicts, on architectures with shared caches.
Suppose two working threads, $T_0$ and $T_1$, are running independently
on the 4-core processor in Figure.~\ref{fig:hierarchy}
to accomplish a GEMM operation.
At time $t_s$, $T_0$ prefetches some data at address $addr_0$
to cache set $\varphi_2(addr_0)$ of the shared L2 cache.
The data will be actually read at time $t_e$.
Then at some time $t_m$ ($t_s < t_m < t_e$),
$T_1$ prefetches data at addr $addr_1$ to the L2 cache.
If $\varphi_2(addr_1)=\varphi_2(addr_0)$, then the
data at $addr_1$ and $addr_0$ will be fetched into the same cache set.
As the shared L2 cache has a non-LRU policy, the
data at $addr_0$ may suffer an eviction before it is actually read,
i.e. a conflict miss occurs.

>>>>

In this blocked algorithm, the loops over the $N$, $K$ and $M$
matrix dimensions are tiled by sizes $N_c$, $K_c$ and $M_c$ at 
layers 1, 2 and 3, respectively.
At layers 4 and 5, the $N$ and $M$ dimensions are further tiled by sizes
$N_r$ and $M_r$, respectively.
As a result, the innermost loop at layer 6 goes over the $K$
dimension for a total of $K_c$ times,
with each iteration performing a rank-1 update on
the $M_r \times N_r$ sub-matrix of $C$, as shown at layer 7.
In OpenBLAS, matrix $C$ is scaled by $\beta$ before the loops,
so the multiply-accumulate operation at layer 7 does not have to scale $C$ again.
Tiling factors $N_c$, $K_c$, $M_c$, $N_r$ and $M_r$ are carefully selected so that
matrices on each level fit into a certain level in the memory hierarchy.
Equations~(\ref{eq:constraints.reg}) -- (\ref{eq:constraints.l3}) express
the constraints on these tiling factors:
\begin{eqnarray}
  es (M_r + N_r + M_r N_r) & \le & c_{0} \label{eq:constraints.reg}\\
  nt_{1} \cdot es (N_r K_c + 2 M_r K_c) & \le & c_{1} \label{eq:constraints.l1}\\
  nt_{2} \cdot es (M_c K_c + 2 N_r K_c) & \le & c_{2} \label{eq:constraints.l2}\\
  es (N_c K_c + nt_{3} \cdot M_c K_c)   & \le & c_{3} \label{eq:constraints.l3}
\end{eqnarray}
Here, $es$ denotes the size of a matrix element,
e.g., 8B for a double precision floating-point number,
and $c_l$ denotes the capacity of a register or cache at level $l$.
By (\ref{eq:constraints.reg}), $M_r$ and $N_r$ are so
constrained that $M_r$ elements from $A$,
$N_r$ elements from $B$ and $M_r \times N_r$ elements from $C$ can
fit into the registers (the pseudo level-0 cache).
By (\ref{eq:constraints.l1}), $K_c$ is so constrained that $B_4$ ($N_r \times K_c$),
$A_4$ ($M_r \times K_c$), and $A_4$'s counterpart in the next iteration
at layer 5 fit into the L1 cache.
By (\ref{eq:constraints.l2}), $M_c$ is so constrained that $A_3$ ($M_c \times K_c$),
$B_3$ ($N_r \times K_c$), and $B_3$'s counterpart in the next iteration
at layer 4 fit into the L2 cache.
Finally, by (\ref{eq:constraints.l3}), $N_c$ is so constrained that 
$B_2$ ($K_c \times N_c$) and $A_2$ ($M_c \times K_c$) fit into the L3 cache.
Larger matrices at higher layers 1 -- 3 always reside in main memory.

@@

As stated in Section~\ref{sec:intro}, a GEMM implementation
is made up by a kernel routine and an overall workload-partitioning strategy.
In Figure~\ref{fig:gemm}, GEBP serves as the kernel routine,
and the overall workload-partitioning strategy is defined by the three outer loops at layers 1 -- 3.
Developers can decide how to partition the workload by choosing
tile factors $M_c$, $N_c$ and $K_c$,
and how to schedule GEBP tasks by interchanging loop orders at layers 1 - 3.
Either the N loop at layer 1 or the M loop at layer 3, or 
both,
can be parallelized in a multi-threaded context.
Specifically, the implementation in Figure~\ref{fig:gemm}
chooses an NKM loop order and parallelizes the M loop at layer 3.
Different threads work on different parts of $A_1$
and share the same $B_1$.

As a result, there are multiple $A_2$ instances (one per thread)
and a single $B_2$ instance (shared by all threads).
To understand how the loop at layer 3 is parallelized on the 4-core processor, 
Figure.~\ref{fig:workload} shows the data accessing patterns of a single thread $T_1$.
While all color shaded submatrices are accessed by $T_1$,
only the z-curve masked submatrices (a subset of color shaded ones)
are packed by the thread.
In other words, each thread packs its own $A_2$ instance and  all the threads collaboratively pack the unique $B_2$ instance.
%% We will use the OpenBLAS GEMM implementation in Figure.~\ref{fig:gemm}
%% in discussion and evaluation throughout the paper.

@@

In contrary to intra-thread cache conflicts,
there exists another kind of cache conflicts, called
inter-thread cache conflicts, on architectures with shared caches.
Suppose two working threads, $T_0$ and $T_1$, are running independently
on the 4-core processor in Figure.~\ref{fig:hierarchy}
to accomplish a GEMM operation.
At time $t_s$, $T_0$ prefetches some data at address $addr_0$
to cache set $\varphi_2(addr_0)$ of the shared L2 cache.
The data will be actually read at time $t_e$.
Then at some time $t_m$ ($t_s < t_m < t_e$),

$T_1$ prefetches data at address $addr_1$ to the L2 cache.
If $\varphi_2(addr_1)=\varphi_2(addr_0)$, then the
data at $addr_1$ and $addr_0$ will be fetched into the same cache set.
As the shared L2 cache has a non-LRU policy,
the data at $addr_0$ may suffer an eviction before it is actually read,
i.e., a conflict miss occurs.

### 2018-05-26 evaluation.tex

In our evaluation, we use the tiling factors
computed as in Section~\ref{subsec:example}:
$M_r = 4$, $N_r = 8$, $K_c = 256$, $M_c = 192$, $N_c = 1024NT$.
The parallelism is controlled by two parameters,
the number of active clusters $NC$
and the number of active threads per cluster $NT_C$.
The total number of threads is $NT = NC \cdot NT_C$.
For the Phytium 2000+ processor, we have
$NC \in [1, 16]$ and $NT_C \in [1, 4]$. 
Generally, DGEMM performance is presented in $flops$.
%% floating-point operations per second ($flops$).
As $NT$ varies in our evaluation,
we use another metric derived from $flops$,
the average thread efficiency,
to describe our performance results consistently.
The average thread efficiency is a normalized value computed as
$E_{avg} = flops / (NT \cdot \widehat{flops})$,
where
$\widehat{flops}$ stands for the theoretical performance peak of a single core.
%% \begin{equation*}
%%   E = flops / (NT \cdot \widehat{flops})
%% \end{equation*}

@@

As the L2 cache is shared by cores inside a cluster,
the impact of cache sharing can be demonstrated
by comparing the DGEMM performance results achieved
with the same $NT$
but different $NT_C$'s.
Figure~\ref{fig:drawback} shows the results.
Each curve represents a parallelism configuration.
Different subfigures are configured with different
$NT$'s,
and different curves in the same subfigure are
configured with different $NT_C$'s.
As there is only one configuration for $NT=64$ ($NC=16$ and $NT_C=4$)
and no comparison is available, the
results for $NT=64$ and $NT=32$ are combined in Figure~\ref{fig:drawback-32}.

@@

We evaluate SCP using the same method as in Section~\ref{subsec:drawback}.
SCP is not evaluated with $NT_C=1$ because there is no need
to partition the cache if $NT_C=1$.
Figure~\ref{fig:benefit} shows the results.
For comparison purposes, the results without SCP
(denoted as Base) are also presented.

@@

Privatization incurs more packing overhead because
redundant packing of $B_2$ is enforced.
Figure~\ref{fig:packb} shows three packing strategies for $B_2$
in a parallelism configuration specified by
$NT_C=4$ and $NC=4$.
The whole packing workload of $B_2$ is divided into 16 tasks
and the packing workload of thread $T_1$ is masked by z-curves. 
%% and the task assignment is annoted in Figure~\ref{fig:packb}.
In the conventional packing approach, each thread takes a single task.
In full privatization, all tasks are replicated to all threads.
As a result, the packing overhead grows in proportion to $NT$
in a full privatization strategy, which is unacceptable
if GEMM is highly parallelized, e.g., when $NT=64$.
Partial privatization in Figure~\ref{fig:packb.partial}
offers an alternative to full privatization.
In partial privatization, privatization only occurs inside
a cluster so that the tasks for one cluster need not to be
replicated to other clusters.
Consequently,  the extra packing overhead of the partial privatization
strategy is bounded by $NT_C$.
For example, on Phytium 2000+, overhead of packing $B_2$
in partial privatization is limited to 4 times
of that in conventional packing.

Figure~\ref{fig:privb} shows the results of the partial
privatization strategy with $NT_C=4$ and $NC=4$.
The results under other parallelism configurations are similar.
For comparison purposes,
SCP with and without privatization are both presented
(denoted as SCP-P and SCP in Figure~\ref{fig:privb}, respectively).
Figure~\ref{fig:privb.papi} shows the cache miss rates.
Privatization achieves slightly lower miss rates for
both L1 and L2 caches.
Figure~\ref{fig:privb.ate} shows the average thread efficiency.
Despite its lower cache miss rates,
privatization suffers a performance loss compared to SCP.
Why is there a performance drop when cache miss rates 
are reduced?
The reason lies in the extra packing overhead introduced by privatization.
Figure~\ref{fig:privb.breakdown} shows the occupancy
ratios of the overall execution times for various overheads.
Overheads in GEMM include
(1) packing of $A_2$, (2) packing of $B_2$,
and (3) synchronization.
While SCP-P involves roughly the same overheads in
packing $A_2$ and synchronization as SCP,
it spends much more time in packing $B_2$.
Table~\ref{tab:breakdown} gives a breakdown of the
execution time, indicating that the overhead for packing $B_2$
is multiplied by a factor of $NT_C=4$,
and the total overhead increases from $2.35\%$ to $4.72\%$.

>>>>

In our evaluation, we use the tiling factors
computed as in Section~\ref{subsec:example}:
$M_r = 4$, $N_r = 8$, $K_c = 256$, $M_c = 192$, $N_c = 1024NT$.
The parallelism is controlled by two parameters,
the number of active clusters $NC$
and the number of active threads per cluster $NT_C$.
The total number of threads is $NT = NC \cdot NT_C$.
For the Phytium 2000+ processor, we have
$NC \in [1, 16]$ and $NT_C \in [1, 4]$. 
Generally, DGEMM performance is presented in $flops$.
As $NT$ varies in our evaluation,
we use another metric derived from $flops$,
the average thread efficiency,
to describe our performance results consistently.
The average thread efficiency is a normalized value computed as
$E_{avg} = flops / (NT \cdot \widehat{flops})$,
where $\widehat{flops}$ stands for the theoretical performance
peak of a single core.
%% \begin{equation*}
%%   E = flops / (NT \cdot \widehat{flops})
%% \end{equation*}

@@

As the L2 cache is shared by cores inside a cluster,
the impact of cache sharing can be demonstrated
by comparing the DGEMM performance results achieved
with the same $NT$
but different $NT_C$'s.
Figure~\ref{fig:drawback} shows the results.
Each curve represents a parallelism configuration.

Different subfigures are configured with different $NT$'s,
and different curves in the same subfigure are
configured with different $NT_C$'s.
As there is only one configuration for $NT=64$ ($NC=16$ and $NT_C=4$)
and no comparison is available,
the results for $NT=64$ and $NT=32$ are combined in Figure~\ref{fig:drawback-32}.

@@

We evaluate SCP using the same method as in Section~\ref{subsec:drawback}.
SCP is not evaluated with $NT_C=1$ because there is no need
to partition the cache if $NT_C=1$.

Figure.~\ref{fig:benefit} shows the results.
For comparison purposes, the results without SCP
(denoted as Base) are also presented.

@@

Privatization incurs more packing overhead because
redundant packing of $B_2$ is enforced.
Figure~\ref{fig:packb} shows three packing strategies for $B_2$
in a parallelism configuration specified by
$NT_C=4$ and $NC=4$.
The whole packing workload of $B_2$ is divided into 16 tasks
and the packing workload of thread $T_1$ is masked by z-curves. 
%% and the task assignment is annoted in Figure~\ref{fig:packb}.
In the conventional packing approach, each thread takes a single task.
In full privatization, all tasks are replicated to all threads.
As a result, the packing overhead grows in proportion to $NT$

in a full privatization strategy, which is unacceptable
if GEMM is highly parallelized, e.g., when $NT=64$.
Partial privatization in Figure.~\ref{fig:packb.partial}
offers an alternative to full privatization.
In partial privatization, privatization only occurs inside
a cluster so that the tasks for one cluster need not to be
replicated to other clusters.
Consequently,  the extra packing overhead of the partial privatization
strategy is bounded by $NT_C$.
For example, on Phytium 2000+, overhead of packing $B_2$
in partial privatization is limited to 4 times
of that in conventional packing.

Figure~\ref{fig:privb} shows the results of the partial
privatization strategy with $NT_C=4$ and $NC=4$.

The results under other parallelism configurations are similar.
For comparison purposes,
SCP with and without privatization are both presented
(denoted as SCP-P and SCP in Figure~\ref{fig:privb}, respectively).
Figure~\ref{fig:privb.papi} shows the cache miss rates.
Privatization achieves slightly lower miss rates for
both L1 and L2 caches.
Figure~\ref{fig:privb.ate} shows the average thread efficiency.
Despite its lower cache miss rates,
privatization suffers a performance loss compared to SCP.
Why is there a performance drop when cache miss rates 
are reduced?
The reason lies in the extra packing overhead introduced by privatization.
Figure~\ref{fig:privb.breakdown} shows the occupancy
ratios of the overall execution times for various overheads.
Overheads in GEMM include
(1) packing of $A_2$, (2) packing of $B_2$,
and (3) synchronization.
While SCP-P involves roughly the same overheads in
packing $A_2$ and synchronization as SCP,
it spends much more time in packing $B_2$.
Table~\ref{tab:breakdown} gives a breakdown of the
execution time, indicating that the overhead for packing $B_2$
is multiplied by a factor of $NT_C=4$,
and the total overhead increases from $2.35\%$ to $4.72\%$.

### 2018-05-26 intro.tex

\section{Introduction}\label{sec:intro}
Dense linear algebra libraries are the most fundamental software in
scientific and engineering computing domains.
Basic Linear Algebra Subprograms (BLAS) (FIXME) defines a collection
of APIs which act as standard building blocks for dense matrix operations.
BLAS routines are divided into 3 levels,
level-1 for vector-vector operations,
level-2 for matrix-vector operations,
and level-3 for matrix-matrix operations.
Processor vendors often provide BLAS implementations
that are highly optimized for their processors,
such as Intel MKL, AMD ACML and NVIDIA cuBLAS.
The HPC community has also contributed several high-quality
BLAS implementations, e.g. ATLAS~\cite{atlas},
GotoBLAS~\cite{gotoblas}, OpenBLAS~\cite{openblas},
and BLIS~\cite{blis,blisport}.

>>>>

\section{Introduction}\label{sec:intro}
Dense linear algebra libraries are the most fundamental software in
scientific and engineering computing domains.
Basic Linear Algebra Subprograms (BLAS) (FIXME) defines a collection
of APIs which act as standard building blocks for dense matrix operations.
BLAS routines are divided into 3 levels,
level-1 for vector-vector operations,
level-2 for matrix-vector operations,
and level-3 for matrix-matrix operations.
Processor vendors often provide BLAS implementations
that are highly optimized for their processors,
such as Intel MKL, AMD ACML and NVIDIA cuBLAS.
The HPC community has also contributed several high-quality
BLAS implementations, e.g., ATLAS~\cite{atlas},
GotoBLAS~\cite{gotoblas}, OpenBLAS~\cite{openblas},
and BLIS~\cite{blis,blisport}.
